Initial estuarine response to the nutrient-rich Piney Point release into Tampa Bay, Florida

### 2022-02-09 revision/review_response.Rmd

```{r setup, echo = F, warning = F, message = F, results = 'hide'}
# figure path, chunk options
knitr::opts_chunk$set(fig.path = 'figs/', warning = F, message = F, echo = F, cache = F, dev.args = list(family = 'serif'), dpi = 300, warning = F)
```

```{r echo = F, cache = F, eval = F}
spelling::spell_check_files(here::here('revision/review_response.Rmd'))
```

We sincerely thank the reviewers for providing comments on our manuscript.  We provide a point-by-point response to these comments below. In general, most of the revisions address concerns raised by all three reviewers, including 1) numerous edits to improve the writing for clarity and brevity, and 2) additional information on sampling methods.  We have also addressed concerns of Reviewer 2 about potential nutrient cycling by adding and expanding a section in the discussion and concerns of Reviewer 4 regarding additional detail on the simulation modelling.

All page and line numbers in the reviewer comments refer to the original submission.  Page numbers noted in our response refer to the revised submission.

@@

P6/L24-39: that is an impressive and extensive list of agencies and partners; any potential drawbacks to so many data streams?

* __Response__: Data quality assurance and control was a primary concern with multiple partners collecting information.  We have included additional information about the data collection procedures and how the Tampa Bay Estuary Program worked to facilitate quality of the data.  Please see the additions starting on page 6.

P6/L54-65: is it routine to take water samples directly from the surface? Ultimately, were the authors worried about any potential stratification, particularly of Karenia brevis.

* __Response__:  Most of the water quality and phytoplankton samples were surface collections following standard sampling methods used by our partners (see the text addition about sampling).  Tampa Bay also has a very shallow average depth of just over 3m and is not strongly stratified for most of the year.  So, we feel confident that the majority of samples used in our analyses were generally representative of conditions in the water column. Anecdotally, benthic fishes were observed during the fish kills (e.g., eels, rays), so it is likely that *K. brevis* was present throughout the water column, although we have very little quantitative data to describe the vertical distribution.  Please see the addition of information on *K. brevis* sampling at the top of page 7 in response to reviewer 1.

@@

P11/L20-: this first part of the results has a choppy structure

* __Response__: Please see the response to the previous comment.

P11/L31-34: the authors indicate the Tampa Bay influx was related to an ongoing coastal bloom; is this indicating the red tide bloom was ongoing prior to the Piney Point discharge? Also, not clear what "related" means in this context. This important statement may require additional context.

* __Response__: Yes, red tide was observed in the Gulf of Mexico prior to the occurrence in Tampa Bay.  Content in the red tide section was edited to make this clear (page 15):

@@

     "This first Tampa Bay influx originated from an ongoing coastal bloom in the Gulf of Mexico, as is common when red tide is observed in the bay [@Steidinger72;@Flaherty11]."

     Blooms in Tampa Bay originate from *K. brevis* that is transported in from the Gulf of Mexico.  An important distinction that we make in this paper is that, although blooms in Tampa Bay are not uncommon, the severity of the 2021 bloom was very likely related to the favorable conditions in the bay prior to July.  These favorable conditions were described in the paper as 1) nutrient availability from Piney Point, and 2) high salinity conditions from low rainfall.  The passage of tropical storm Elsa was also a confounding factor.  It is nearly impossible to identify a single factor causing the observed conditions, but we describe the considerations in the discussion (page 23).

@@

This is an interesting study in that it utilizes a quasi-Lagrangian approach to quantify the influence of a nutrient-rich waste plume on an estuarine ecosystem and is the first that I am aware of that documents the impacts of the Piney Point episode. Nonetheless, there are some points that require clarification to give the reader confidence in the findings. In general, more details are needed on several aspects of the methodology, and I provide specific details below on this as well as other issues.

* __Response__: Thank you for your constructive comments.  Please see the detailed responses below.

Page 1, Line 38 - "Elevated levels of phytoplankton" - needs to be clearer… is this referring to biomass or abundance?

* __Response__: This line was changed to "An initial phytoplankton bloom (non-harmful diatoms) was first observed...".  This was based on chlorophyll measurements in the vicinity of Port Manatee and cell concentrations enumerated to taxa from laboratory analysis.  The word limit for the abstract prevents these details, but they are presented in text (page 18).

@@

     More detailed model description or analysis is beyond the scope this work.  A dedicated manuscript is being prepared to document the TBCOM tracer model as a rapid response to the Piney Point event. Preliminary results can be found in CERF and AGU presentations:

     Liu, Y., Weisberg, R.H., Zheng, L., Sun, Y., Chen, J. (2021), Nowcast/Forecast of the Tampa Bay, Piney Point Effluent Plume: A Rapid Response, Abstract (OS35B-1036) presented at AGU Fall Meeting, New Orleans, Louisiana, December 2021.

     Liu, Y., Weisberg, R.H., Zheng, L., Sun, Y., Chen, J. (2021), Nowcast/Forecast of the Tampa Bay, Piney Point Effluent Plume: A Rapid Response, Coastal and Estuarine Research Federation (CERF) 2021 Virtual Meeting, November 2021.

     The main text was edited to include some of the above details (page 5): "The modeled plume evolution web product (http://ocgweb.marine.usf.edu/~liu/Tracer/) served as the principal guidance for coordinating the data collection during the event. Preliminary model results for Piney Point are reported in @Liu21 and previous model veracity testing was described in @Chen18 and @Chen19 (and references therein)"

@@

Page 6, Lines 46-54 - More information is needed on laboratory procedures. In particular, with many different entities running samples and presumably utilizing different analytical approaches for each variable, it is important to know how comparable the data is between programs.

* __Response__: Our initial submission included supplement material that described the sample collection and processing in more detail.  We have moved that content from the supplement to the main text, beginning on page 6:

@@

Page 14, Macroalgae and seagrass trends section - This appears to be one of the weaker components of the project, although it is outside of my expertise. My biggest concern is that it is not clear to me how useful "% occurrence" is. As far as I can tell, this is not equivalent to biomass or areal coverage. Perhaps I am mistaken? Regardless, it would be useful to see a better justification for this metric. Also, how did the 2021 data compare to historic spatial-temporal trends?

* __Response__: Frequency occurrence is a routine reporting metric used by the Tampa Bay Estuary Program to summarize the long-term annual transect monitoring [see @Johansson16;@Sherwood17]. Additional text on page 10 was added to make this clear:

     "Frequency occurrence estimates were used to evaluate macroalgae and seagrasses as a standard metric used in previous analyses in Tampa Bay [@Johansson16;@Sherwood17]."

     We realize that additional metrics could have been used to summarize the data and we presented trend assessments using the Braun-Blanquet abundance estimates as confirmation for the results from the frequency occurrence analyses (Tables S3 and S4).  However, this does not provide an assessment of biomass or coverage.  Weight data were collected for select macroalgae samples during the 2021 surveys and a preliminary analysis (see the Seagrass -> Biomass estimates tab [here](https://tbep-tech.github.io/piney-point-analysis/trend)) suggested similar trends as the frequency occurrence results.  Because we did not have comparable weight data for seagrasses, the analysis was not included.  It is also worth mentioning that coverage estimates are obtained every two years for Tampa Bay as an annual snapshot (noted at the top of page 21 in the discussion), but these are not of sufficient temporal resolution to support the analyses in this paper.

     There were limitations to our analysis that were noted in the original draft, primarily the lack of long-term historical data on seasonal trends for seagrasses and macroalgae.  For example, page 22,line 52 in the original discussion: "Long-term monitoring data describing normal seasonal variation in macroalgae are unavailable and we cannot distinguish between seasonal and interannual changes and those in potential response to the Piney Point release."

@@

Discussion - In general, it would be useful if the team would spend some time establishing the prevalence of N-limitation in the system, either based on their own data or previous studies. The interpretation of plume impacts is predicated on the system being N-limited, so that needs to be established up front.

* __Response__: We agree that the nutrient limitation needs to be clearly stated as the basis for focusing primarily on nitrogen and not phosphorus. The second paragraph in the introduction was meant to provide this justification, but this was clearly insufficient as noted in the review.  We include some additional text and citations in the final paragraph of the introduction as further justification - see the response to the similar critique from reviewer 2.  Also note that phosphorus summary plots were previously included in the supplement.

Page 20, Line 22 - Ammonium, not ammonia. Or was pH such that ammonia would be the dominant form?

* __Response__: This is in reference to total ammonia nitrogen, as was provided by the water quality sampling.  We have made a note in the methods section (top of page 6) that any mention of ammonia refers to total ammonia nitrogen.

@@

Page 20, Line 39 - Might be worth having a separate paragraph that talks about the timing/duration of the K. brevis bloom in 2021 compared to historic conditions.

* __Response__: Previous content in the discussion below this paragraph provided some historical context on red tide events in the region.  Specifically, on page 23: "Occurrence of this species has historically been spatially distinct, with blooms originating in subsurface water offshore on the West Florida Shelf..."

### 2022-02-09 manu-draft.Rmd
formatting done for revision resubmission to MPB
revision submitted to MPB

Legacy mining facilities pose significant risks to aquatic resources.  From March 30th to April 9th, 2021, 814 million liters of phosphate mining wastewater and marine dredge water from the Piney Point facility were released into lower Tampa Bay (Florida, USA). This resulted in an estimated addition of 186 metric tons of total nitrogen, exceeding typical annual external nitrogen load estimates to lower Tampa Bay in a matter of days. An initial phytoplankton bloom (non-harmful diatoms) was first observed in April. Filamentous cyanobacteria blooms (*Dapis* spp.) peaked in June, followed by a bloom of the red tide organism *Karenia brevis*. Reported fish kills tracked *K. brevis* concentrations, prompting cleanup of over 1600 metric tons of dead fish. Seagrasses had minimal changes over the study period.  By comparing these results to baseline environmental monitoring data, we demonstrate adverse water quality changes in response to abnormally high and rapidly delivered nitrogen loads into Tampa Bay.

@@

## Data Analysis

@@

Several of the water quality responses are consistent with observations of nutrient loading in other shallow Gulf Coast estuaries [@Doering06;@Caffrey14;@Greening14].  The relationship between nutrients, chl-a, and water transparency followed expectations of reduced water quality with increased nutrient loads. Temporally, these changes were observed at different times and for different species of phytoplankton.  The initial increase in chl-a was first associated with a diatom bloom in April.  The red tide species *K. brevis* was also first introduced to Tampa Bay from the Gulf of Mexico in April, but was not observed at high densities in the Bay until June and July. Peaks in dissolved oxygen saturation were also observed as an indicator of elevated phytoplankton production [@Kemp80], particularly in July with the peak *K. brevis* bloom (Figures S2, S3).  Of note is that inorganic species of nitrogen, mainly ammonia, were only present at high concentrations in early April.  Management concerns of the negative impacts of nutrients on water quality focused primarily on the high concentrations of ammonia in the discharge (Table \@ref(tab:stktab)), which can be utilized rapidly by many phytoplankton taxa [@Bates76;@Domingues11].  Low concentrations of ammonia after April may be explained by quick uptake by the initial diatom bloom, where TN that included particulate and dissolved organic sources was at high concentrations through April and again peaked in July.  Variation in observed concentrations of nutrients is complex given that high concentrations may suggest availability to support phytoplankton growth, whereas low concentrations may imply cycling of available nitrogen in organic forms already utilized by different taxa, including macroalgae [@Valiela97;@Cohen06].

>>>>

Legacy mining facilities pose significant risks to aquatic resources.  From March 30th to April 9th, 2021, 814 million liters of phosphate mining wastewater and marine dredge water from the Piney Point facility were released into lower Tampa Bay (Florida, USA). This resulted in an estimated addition of 186 metric tons of total nitrogen, exceeding typical annual external nitrogen load estimates to lower Tampa Bay in a matter of days. An initial phytoplankton bloom (non-harmful diatoms) was first observed in April. Filamentous cyanobacteria blooms (*Dapis* spp.) peaked in June, followed by a bloom of the red tide organism *Karenia brevis*. Reported fish kills tracked *K. brevis* concentrations, prompting cleanup of over 1600 metric tons of dead fish. Seagrasses had minimal changes over the study period.  By comparing these results to baseline environmental monitoring data, we demonstrate adverse water quality changes in response to abnormally high and rapidly delivered nitrogen loads.

@@

## Data analysis

@@

Several of the water quality responses are consistent with observations of nutrient loading in other shallow Gulf Coast estuaries [@Doering06;@Caffrey14;@Greening14].  The relationship between nutrients, chl-a, and water transparency followed expectations of reduced water quality with increased nutrient loads. Temporally, these changes were observed at different times and for different species of phytoplankton.  The initial increase in chl-a was first associated with a diatom bloom in April.  The red tide species *K. brevis* was also first introduced to Tampa Bay from the Gulf of Mexico in April, but was not observed at high densities in the Bay until June and July. Peaks in dissolved oxygen saturation were also observed as an indicator of elevated phytoplankton production [@Kemp80], particularly in July with the peak *K. brevis* bloom (Figures S2d, S3d).  Of note is that inorganic species of nitrogen, mainly ammonia, were only present at high concentrations in early April.  Management concerns of the negative impacts of nutrients on water quality focused primarily on the high concentrations of ammonia in the discharge (Table \@ref(tab:stktab)), which can be utilized rapidly by many phytoplankton taxa [@Bates76;@Domingues11].  Low concentrations of ammonia after April may be explained by quick uptake by the initial diatom bloom, where TN that included particulate and dissolved organic sources was at high concentrations through April and again peaked in July.  Variation in observed concentrations of nutrients is complex given that high concentrations may suggest availability to support phytoplankton growth, whereas low concentrations may imply cycling of available nitrogen in organic forms already utilized by different taxa, including macroalgae [@Valiela97;@Cohen06].
